---
title: "Landscape of Change Results" 
author: "Kyle A. Lima, Catherine Schmitt, Seth Benz, and Peter Nelson"
date: "February, 2022"
header-includes:
    - \usepackage{setspace}\doublespacing
output:
  html_document:
    theme: readable
---

# Have Changes Occured Between 1880 and Now? {.tabset .tabset-fade .tabset-pills}
## Birds

```{r setup, include=FALSE}
require(knitr)
require(utils)
require(dplyr)
require(ggplot2)
library(ggmap)
library(tidyverse)

freq.table <- read.csv('frequency_table_data.csv', header = TRUE)
freq.plot <- read.csv('frequency_plot_data.csv', header = TRUE)
modern.table <- read.csv('modern_table_data.csv', header = TRUE)
modern <- read.csv('modern_fulltable_data.csv', header = TRUE)
dec.table <- read.csv('drasticdec_table_data.csv', header = TRUE)
inc.table <- read.csv('drasticinc_table_data.csv', header = TRUE)
bird.mod.locs <- read.csv('species_plotting_data.csv', header = TRUE)

#Make separate data.frames for each species that is decreasing
AMWO <- bird.mod.locs %>% 
  filter(bird.mod.locs$common.name=='American Woodcock')
 
YBFL <- bird.mod.locs %>% 
  filter(bird.mod.locs$common.name=='Yellow-bellied Flycatcher')

BCNH <- bird.mod.locs %>% 
  filter(bird.mod.locs$common.name=='Black-crowned Night-Heron')

#Make separate data.frames for each species that is increasing
BTBW <- bird.mod.locs %>% 
  filter(bird.mod.locs$common.name=='Black-throated Blue Warbler')

EAPH <- bird.mod.locs %>% 
  filter(bird.mod.locs$common.name=='Eastern Phoebe')

BLJA <- bird.mod.locs %>% 
  filter(bird.mod.locs$common.name=='Blue Jay')

#MDI Map for species plots
mdi.map <- get_stamenmap(
  bbox = c(left = -68.5, bottom = 44.17, right = -68.13, top = 44.5),
  maptype = 'terrain',
  zoom = 11)

```



### Summary

How have bird species populations changed from 1880 to now? Bird species abundance and diversity has certainly changed between now and the 1880's when the Champlain Society was running field courses on Mount Desert Island (MDI). We compared the Champlain Society's species list that they kept over the 4 year duration they were active on MDI, to the past 4 years of eBird and iNaturalist data from 2018-2021. Today, there are `r length(which(modern$frequency.y=='common'))` common species on MDI, whereas in the early 1880's there were only `r length(which(modern$frequency.x=='common'))` common species. So, the diveristy of regularly occuring species has increased. **Table 1** lists all species that were not recorded historically or were rare, but are now common or have increased greatly since the 1880s. When looking at the changes of the species recorded historically we do see a `r (round(((length(which(freq.table$frequency.changes=='decreased'))*100)/88), digits=2))`% decrease in these species (**Fiugre 1**, **Table 3**). **Table 1** and **Table 2** highlight the species with noteable changes over the last 150 years. For a full list of changes to species see **Table 4**.

<br>

### Bar Graph of Species Changes

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 1**. The bar graph shows changes in the species recorded by the Champlain Society in the 1880s.'}

freq.plot %>% 
  ggplot(aes(freq.changes, species.number, fill = freq.changes)) +
  geom_bar(stat="identity", color="black") +
  ggtitle("Species Frequency Changes from 1880 to Present") + 
  geom_text(aes(label = species.number, vjust = 2)) +
  guides(fill = 'none') +
  labs(x = "Change in Frequency", y = "Number of Species") +
  theme_classic(base_size=14) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("forestgreen", "darkslategray4", "darkorange3"))

```

<br>

### Notably Increasing Species

**Table 1**. This table shows the species that have increased drastically in abundance since the 1880s. From this, we can infer that these species have expanded their range likely as a result of cliamte change and landscape change. There are `r length(which(modern.table$frequency=='common'))` species that are now common on MDI that were not present historically, and `r length(which(modern.table$frequency=='uncommon'))` species that are uncommon but regular and were not present historically.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(inc.table)

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 5**. Black-throated Blue Warbler (*Setophaga caerulescens*) distribution across MDI.'}

#BTBW
ggmap(mdi.map) +
  geom_point(data=BTBW,
             aes(x=longitude, y=latitude),
             shape = 21,
             size = 2.5,
             color = 'black', 
             fill = '#006699',
             stroke = 1,
             alpha = 0.6) +
  theme_classic(base_size = 14 ) +
  ggtitle("Black-throated Blue Warbler Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "longitude", y = "latitude")

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 6**. Blue Jay (*Cyanocitta cristata*) distribution across MDI.'}

#BLJA
ggmap(mdi.map) +
  geom_point(data=BLJA,
             aes(x=longitude, y=latitude),
             shape = 21,
             size = 2.5,
             color = 'black',
             fill = '#0099FF',
             stroke = 1,
             alpha = 0.5) +
  theme_classic(base_size = 14 ) +
  ggtitle("Blue Jay Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "longitude", y = "latitude")

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 7**. Eastern Phoebe (*Sayornis phoebe*) distribution across MDI.'}

#EAPH
ggmap(mdi.map) +
  geom_point(data=EAPH,
             aes(x=longitude, y=latitude),
             shape = 21,
             size = 2.5,
             color = 'black', 
             fill = '#996633',
             stroke = 1,
             alpha = 0.6) +
  theme_classic(base_size = 14 ) +
  ggtitle("Eastern Phoebe Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "longitude", y = "latitude")

```


<br>

### Noteably Decreasing Species

**Table 2**. This table shows the species that have decreased drastically in abundance since the 1880s.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(dec.table)

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 2**. American Woodcock (*Scolopax minor*) distribution across MDI.'}

#AMWO
ggmap(mdi.map) +
  geom_point(data=AMWO,
             aes(x=longitude, y=latitude),
             shape = 21,
             size = 2.5,
             color = 'black', 
             fill = '#FF9933',
             stroke = 1,
             alpha = 0.6) +
  theme_classic(base_size = 14 ) +
  ggtitle("American Woodcock Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "longitude", y = "latitude")

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 3**. Black-crowned Night-Heron (*Nycticorax nycticorax*) distribution across MDI.'}

#BCNH
ggmap(mdi.map) +
  geom_point(data=BCNH,
             aes(x=longitude, y=latitude),
             shape = 21,
             size = 2.5,
             color = 'black', 
             fill = '#000000',
             stroke = 1,
             alpha = 0.6) +
  theme_classic(base_size = 14 ) +
  ggtitle("Black-crowned Night-Heron Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "longitude", y = "latitude")

```
<br>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='**Figure 4**. Yellow-bellied Flycatcher (*Empidonax flaviventris*) distribution across MDI.'}

#YBFL
ggmap(mdi.map) +
  geom_point(data=YBFL,
             aes(x=longitude, y=latitude),
             shape = 21,
             size = 2.5,
             color = 'black', 
             fill = '#FFFF66',
             stroke = 1,
             alpha = 0.6) +
  theme_classic(base_size = 14 ) +
  ggtitle("Yellow-bellied Flycatcher Distribution") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "longitude", y = "latitude")

```

<br>

### All Decreasing Species

**Table 3**. This table shows the species we know to have descreased in abundance since the 1880s.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

freq.table.dec <- freq.table %>% 
  filter(freq.table$frequency.changes == 'decreased')

knitr::kable(freq.table.dec)

```

<br>

### Total Species List Changes

**Table 5**. This table breaks down of all species recorded by the Champlain Society with their frquency status from the 1880s and today. If you look at the 'Changes in Frequency' column you can see how these species have changed from historic times.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(freq.table)

```

<br>

### Important Information

The historical dataset from the Champlain Society was collected from early July through the first week of September 1880-1883. To properly compare these data to modern day, we used eBird and iNaturalist data filtered to the same early July through the first week of Septemeber time frame, and to the same number of years (4) from 2018-2021. To compare how often these species were/are detected, we had to create a qualitative descriptor properly assinged to the species as the historical dataset did not allow for a quantitative analysis. These designations were assigned based on the Champlain Society's notes, and other sources describing bird abundance around that time. For the modern dataset we used a count of the number of times a species was reported and designated a category based on these bins: < 5 = 'very rare', 5 > 20 = 'rare', 20 > 50 = 'uncommon', > 50 = 'common.'
  
During the analysis, we compared these detection frequencies across modern and historic times, and separated into their own tables (the 'Notably' increasing or decreasing tables) species with changes greater than one category less common or more common. So, for example if a species was common in the 1880s but is now uncommon, that would not fall into the notable changes. Whereas, if a species was common in the 1880s but is now rare, that would qualify as a noteably decreasing species. Considering the qualitative nature of our data with broad bins in which species were assigned, this makes it very difficult to distinguish fine scale details. So the larger the change the more confident we can be in saying that the species trend we are observing is true.










## Insects
