---
title: "Landscape of Change Results"
output: html_document
---

```{r setup, include=FALSE}
require(knitr)
require(utils)
require(dplyr)
require(ggplot2)

freq.table <- read.csv('frequency_table_data.csv', header = TRUE)
freq.plot <- read.csv('frequency_plot_data.csv', header = TRUE)
modern.table <- read.csv('modern_table_data.csv', header = TRUE)
modern <- read.csv('modern_fulltable_data.csv', header = TRUE)
```

## How have bird species populations changed from 1880 to now?

### Summary

Bird species abundance and diversity has certainly changed between now and the 1880's when the Champlain Society was running field courses on Mount Desert Island (MDI). We compared the Champlain Society's species list that they kept over the 4 year duration they were active on MDI, to the past 4 years of eBird data from 2018-2021. Since the effort of birders today is much more extensive than what could have been possible in the 1880's lets compare the species diversity of common species. Today, there are `r length(which(modern$frequency.y=='common'))` common species on MDI, whereas in the early 1880's there were only `r length(which(modern$frequency.x=='common'))` common species. So, the diveristy of regularly occuring species has increased. **Table 3** lists all species that were not recorded historically, but are now common. When looking at the changes of the species recorded historically we do see a drastic `r (length(which(freq.table$frequency.changes=='decreased'))*100)/88`% decrease in these species (**Fiugre 1**, **Table 1**). For a full list of changes to species see **Table 2**.

### Bar Graph of Species Changes

```{r, echo=FALSE, message=FALSE, warning=FALSE}

freq.plot %>% 
  ggplot(aes(freq.changes, species.number, fill = freq.changes)) +
  geom_bar(stat="identity", color="black") +
  ggtitle("Species Frequency Changes from 1880 to Present") + 
  geom_text(aes(label = species.number, vjust = 2)) +
  guides(fill = FALSE) +
  labs(x = "Change in Frequency", y = "Number of Species") +
  theme_classic(base_size=14) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values=c("forestgreen", "darkslategray4", "darkorange3"))

```

**Figure 1**. The bar graph shows changes in the species recorded by the Champlain Society in the 1880s.


### Species That Decreased


**Table 1**. This table shows the species we know to have descreased in abundance since the 1880s. The frequency value 'migrant only' refers to the species not breeding on MDI and isn't specifically associated with migratory species.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

freq.table.dec <- freq.table %>% 
  filter(freq.table$frequency.changes == 'decreased')

knitr::kable(freq.table.dec)

```


### Total Species List Changes


**Table 2**. This table breaks down of all species recorded by the Champlain Society with their frquency status from the 1880s and today. If you look at the 'Changes in Frequency' column you can see how these species have changed from historic times.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(freq.table)

```


### Modern Species Not Recorded in the 1880s


**Table 3**. This table shows the common and uncommon species that are reguarly seen on MDI in modern times, but were not recorded in the 1880s. From this, we can infer that these species have expanded their range likely as a result of cliamte change and landscape change. There are `r length(which(modern.table$frequency=='common'))` species that are now common on MDI that were not present historically, and `r length(which(modern.table$frequency=='uncommon'))` species that are uncommon but regular and were not present historically.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::kable(modern.table)

```

